Affective Uplift During Video Game Play: A Naturalistic Case Study

### 2023-10-19 ms.Rmd
Remove Quarto
Submission formatting

wordcount: '`r scales::number(wordcountaddin:::word_count("ms.Rmd"), big.mark = ",")`'
bibliography: references.bib
keywords: "video games, well-being, mood, play behavior, telemetry"
csl: apa.csl
floatsintext: yes
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
output: 
  papaja::apa6_pdf:
    number_sections: false
    keep_tex: true
  papaja::apa6_docx: 
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

# Introduction

@@

Understanding whether and when games' short-term effects emerge is critical for establishing games' potential for mood-related interventions, as well as for building a theoretical foundation for repeated short-term gaming experiences' long-term effects on mental health. At present, however, the prevalence and magnitude of these short-term effects remains poorly understood. Some existing evidence suggests that games can provide short-term boosts to well-being ["mood repair"; @bowmanMoodGameSelective2015; @tyackRestorativePlayVideogames2020], possibly to a greater extent than non-interactive media such as videos [@riegerEatingGhostsUnderlying2015]. On the other hand, games might affect players negatively: Frustrating gaming experiences, for example, can lead to negative consequences such as immediate post-play aggression [@przybylskiCompetenceimpedingElectronicGames2014].

@@

Here, we aim to address these three challenges to better understand how real play in natural contexts might predict mood on short timescales. Specifically, we examined an intensive longitudinal dataset from the popular commercially available game PowerWash Simulator [PWS; @vuorreIntensiveLongitudinalDataset2023], which includes mood questions embedded in the game itself, to ask three questions: First, to what extent does mood change from immediately before video game play to during play? Second, hyow heterogeneous are these changes in the population of similar players? And third, how do changes in mood develop over the course of a gaming session?

@@

# Write data for any supplementary notebooks
dir.create("data", FALSE)
write_rds("data/data.rds")
```

@@

We measured mood with a single item: "How are you feeling right now?" [@killingsworthWanderingMindUnhappy2010]. Participants responded using a visual analogue scale (VAS) with endpoints "Very bad" and "Very good" that recorded 1000 possible values, which we rescaled to the unit interval (0-1) for this study. Consequently, our results can also be interpreted on the "[proportion] of maximum possible" scale [POMP; @cohenProblemUnitsCircumstance1999]. While well-being is often studied with multi-item scales to differentiate between dimensions of positive and negative affect, the frequent probing of mental states in this study required a minimally intrusive instrument that would interrupt the participants' play experience as little as possible. Moreover, such single-item assessments have previously been validated and are recommended for intensive longitudinal studies [@songExaminingConcurrentPredictive2022].

@@

```{r}
#| label: fig-ct1
#| include: true
#| fig.cap: "Estimated (changes) in mood as a function of session duration. Top: Average mood during a gaming session. Bottom: Change in mood during a session compared to mood at the beginning of a session. Gray ribbons indicate 95\\% confidence bands. We truncated the x-axis at three hours for this figure."
#| fig-height: 3
#| fig-width: 5

@@

Although the overall magnitude of the estimated change was small (`r filter(fit_tbl, variable=="Difference")$res`), it nevertheless is comparable to mood shifts following other frequent voluntary activities. For example, experience sampling studies of US [@killingsworthWanderingMindUnhappy2010] and Korean [@choiTakingStockHappiness2017] adults indicated that watching television (+2% change compared to an individual's average across all activities), reading (+2%), and shopping (+3%) were associated with smaller mood shifts than we observed in association with PWS. On the other hand, those studies indicated that listening to music (+7–9%), eating or cooking (+8%), taking a walk (+9%), visiting an urban green space [+9–10%; @tostNeuralCorrelatesIndividual2019], exercising (+11%), dating (26.5%), sexual intercourse (+28%), or taking a trip (+30.5%) correlate with greater shifts in well-being.

@@

Current theories provide a few candidates regarding mood change mechanisms. Mood management theory [MMT; @zillmannMoodManagementCommunication1988] posits that players choose specific entertainment media as part of a (conscious or subconscious) strategy to regulate valence and arousal and to increase hedonic well-being. Specifically, MMT predicts that when players feel more negative emotions, they are more likely to select games with positive hedonic valence (i.e., games that are generally positive in tone), and that when players are in high arousal states such as anger stress, they are more likely to choose calm and relaxing games with.

@@

As always, the results here are likely to generalize in some ways but not in others. Our sample is somewhat representative with regard to gender and covers a wide age range roughly in line with the age demographics of general US adult video game players [@engelstatterVideoGamesBecome2022], suggesting that our results may generalize to other adult PWS players from Western countries. They are less likely to generalize to younger players and those from non-Western countries. The sample may further suffer from self-selection bias: Given that players voluntarily chose to download the PWS research edition, it is plausible that people who felt more positively toward the game (and more positively while playing it) were more likely to opt into the study. Finally, we studied just one commercial video game with a feature set that is very different from today's most commonly-played PC games [which at the time of writing include, for example, Minecraft, Fortnite, and Baldur's Gate 3, but only one other simulation game, The Sims 4; @newzooMostPopularPC2023].

@@

## Author notes

### Data and Material Availability

### Author contributions

Conceptualization: MV, AKP\
Data Curation: MV, KM\
Formal Analysis: MV, KM\
Funding Acquisition: AKP, MV, KM\
Methodology: MV, KM\
Project Administration: MV, AKP\
Resources: AKP, MV\
Supervision: AKP, MV\
Validation: KM, MV\
Visualization: MV, KM\
Writing – Original Draft: NB, MV, TH, AKP, KM\
Writing – Review & Editing: MV, NB, AKP, KM\

### Declaration of Conflicting Interests

### Funding

@@

>>>>

wordcount: '`r scales::number(wordcountaddin:::word_count("ms.Rmd"), big.mark = ",")`'
bibliography: references.bib
keywords: "video games, well-being, mood, play behavior, telemetry"
csl: apa.csl
floatsintext: yes
linenumbers: no
draft: no
mask: yes
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: doc
output: 
  papaja::apa6_pdf:
    number_sections: false
    keep_tex: true
  papaja::apa6_docx: 
    number_sections: false
editor_options: 
  chunk_output_type: console
---

@@

@@

Understanding whether and when games' short-term effects emerge is critical for establishing games' potential for mood-related interventions, as well as for building a theoretical foundation for repeated short-term gaming experiences' long-term effects on mental health. At present, however, the prevalence and magnitude of these short-term effects remains poorly understood. Some existing evidence suggests that games can provide short-term boosts to well-being ["mood repair", @bowmanMoodGameSelective2015; @tyackRestorativePlayVideogames2020], possibly to a greater extent than non-interactive media such as videos [@riegerEatingGhostsUnderlying2015]. On the other hand, games might affect players negatively: Frustrating gaming experiences, for example, can lead to negative consequences such as immediate post-play aggression [@przybylskiCompetenceimpedingElectronicGames2014].

@@

Here, we aim to address these three challenges to better understand how real play in natural contexts might predict mood on short timescales. Specifically, we examined an intensive longitudinal dataset from the popular commercially available game PowerWash Simulator [PWS, @vuorreIntensiveLongitudinalDataset2023], which includes mood questions embedded in the game itself, to ask three questions: First, to what extent does mood change from immediately before video game play to during play? Second, hyow heterogeneous are these changes in the population of similar players? And third, how do changes in mood develop over the course of a gaming session?

@@

# Write data for any supplementary notebooks
dir.create("data", FALSE)
write_rds(dat, "data/data.rds")
```

@@

We measured mood with a single item: "How are you feeling right now?" [@killingsworthWanderingMindUnhappy2010]. Participants responded using a visual analogue scale (VAS) with endpoints "Very bad" and "Very good" that recorded 1000 possible values, which we rescaled to the unit interval (0-1) for this study. Consequently, our results can also be interpreted on the "[proportion] of maximum possible" scale [POMP, @cohenProblemUnitsCircumstance1999]. While well-being is often studied with multi-item scales to differentiate between dimensions of positive and negative affect, the frequent probing of mental states in this study required a minimally intrusive instrument that would interrupt the participants' play experience as little as possible. Moreover, such single-item assessments have previously been validated and are recommended for intensive longitudinal studies [@songExaminingConcurrentPredictive2022].

@@

```{r}
#| label: fig-ct1
#| include: true
#| fig.cap: "Estimated (changes) in mood as a function of session duration. Top: Average mood during a gaming session. Bottom: Change in mood during a session compared to mood at the beginning of a session. Gray ribbons indicate 95\\% confidence bands. We truncated the x-axis at three hours for this figure."
#| fig-height: 3
#| fig-width: 8

@@

Although the overall magnitude of the estimated change was small (`r filter(fit_tbl, variable=="Difference")$res`), it nevertheless is comparable to mood shifts following other frequent voluntary activities. For example, experience sampling studies of US [@killingsworthWanderingMindUnhappy2010] and Korean [@choiTakingStockHappiness2017] adults indicated that watching television (+2% change compared to an individual's average across all activities), reading (+2%), and shopping (+3%) were associated with smaller mood shifts than we observed in association with PWS. On the other hand, those studies indicated that listening to music (+7–9%), eating or cooking (+8%), taking a walk (+9%), visiting an urban green space [+9–10%, @tostNeuralCorrelatesIndividual2019], exercising (+11%), dating (26.5%), sexual intercourse (+28%), or taking a trip (+30.5%) correlate with greater shifts in well-being.

@@

Current theories provide a few candidates regarding mood change mechanisms. Mood management theory [MMT, @zillmannMoodManagementCommunication1988] posits that players choose specific entertainment media as part of a (conscious or subconscious) strategy to regulate valence and arousal and to increase hedonic well-being. Specifically, MMT predicts that when players feel more negative emotions, they are more likely to select games with positive hedonic valence (i.e., games that are generally positive in tone), and that when players are in high arousal states such as anger stress, they are more likely to choose calm and relaxing games with.

@@

As always, the results here are likely to generalize in some ways but not in others. Our sample is somewhat representative with regard to gender and covers a wide age range roughly in line with the age demographics of general US adult video game players [@engelstatterVideoGamesBecome2022], suggesting that our results may generalize to other adult PWS players from Western countries. They are less likely to generalize to younger players and those from non-Western countries. The sample may further suffer from self-selection bias: Given that players voluntarily chose to download the PWS research edition, it is plausible that people who felt more positively toward the game (and more positively while playing it) were more likely to opt into the study. Finally, we studied just one commercial video game with a feature set that is very different from today's most commonly-played PC games [which at the time of writing include, for example, Minecraft, Fortnite, and Baldur's Gate 3, but only one other simulation game, The Sims 4, @newzooMostPopularPC2023].

@@

@@

# Contributions

\noindent
Conceptualization: MV, AKP\
Data Curation: MV, KM\
Formal Analysis: MV, KM\
Funding Acquisition: AKP, MV, KM\
Methodology: MV, KM\
Project Administration: MV, AKP\
Resources: AKP, MV\
Supervision: AKP, MV\
Validation: KM, MV\
Visualization: MV, KM\
Writing – Original Draft: NB, MV, TH, AKP, KM\
Writing – Review & Editing: MV, NB, TH, AKP, KM\

# Funding information

# Competing interests

# Data accessibility statement

# Figure titles and legends

## Figure 1.

A. Scatterplots of three participants' (rows) mood responses (pre-play: red; during play: blue) over eight sessions' (columns) durations. B. Histograms of session-mean (C. person-mean) moods before (top) and during (bottom) play sessions. D. Differences in session-mean (E. player-mean) mood differences (during session - pre-play). F. Scatterplot of person-mean mood reports at the beginning (x-axis) and during gameplay sessions (y-axis). Identity line is shown in green, and an exploratory GAM regression line is shown in blue.

## Figure 2.

Estimated (changes) in mood as a function of session duration. Top: Average mood during a gaming session. Bottom: Change in mood during a session compared to mood at the beginning of a session. Gray ribbons indicate 95\\% confidence bands. We truncated the x-axis at three hours for this figure.

## Figure 3.

Estimated (changes) in mood as a function of session duration and pre-play mood. Top. Mood for the average player during a gaming session with pre-play mood at 5th, 25th, 50th, and 75th percentiles (columns). Ribbons indicate 95\\% confidence. Bottom. Same as above but with change in mood on the y-axis.
